---
title: "retail"
output: word_document
author: "Joyce Woznica"
date: "October 19,2019"
---

This is a tutorial uses R package "arules" to analyze the market basket data. 

# Load the libraries
```{r}
# install.packages(arules)
# install.packages(arulesViz)
library(arules)
library(arulesViz)
```

# Load the dataset
```{r}
retail <- read.transactions("C:/Users/Joyce/Desktop/Syracuse/IST707/Classwork/week3/retail.csv", format="basket", sep=",")
```

# We can explore the data before we make any rules. The following script is going to create an item frequency plot for the top 20 items.
# "type" indicates whether item frequencies should be displayed relative or absolute
```{r}
itemFrequencyPlot(retail,topN=20,type="absolute")
```

# Mine rules with the Association Rule algorithm. 
# It is required to set the minimum support and confidence values.
```{r}
rules <- apriori(retail, parameter = list(supp = 0.001, conf = 0.8,maxlen=3))
```

# Show the top 5 rules, rounding with 2 digits
# From the top 5 rules, we could notice that the rule {bottled beer, soups} => {whole milk} is strong, indicating bottled beer and soups are frequently bought together with whole milk. So we would suggest to place these items close to each other.
```{r}
options(digits=2)
inspect(rules[1:5])
```

# Get summary info about all rules
```{r}
summary(rules)
```

# Sort rules so that we can view the most relevant rules first. For example, sort rules with "confidence":
```{r}
rules<-sort(rules, by="confidence", decreasing=TRUE)
subset.matrix <- is.subset(rules, rules)
subset.matrix[lower.tri(subset.matrix, diag=T)] <- NA
redundant <- colSums(subset.matrix, na.rm=T) >= 1
rules.pruned <- rules[!redundant]
rules<-rules.pruned
```

# Visualize the rules
```{r}
#plot(rules,method="graph",interactive=TRUE,shading=NA)
```